---
title: "tzLookup"
author: "Mark Hagemann"
date: "May 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm implementing a timezone lookup for arbitrary locations within the US. The data I have to work with looks like this: 

```{r, echo = FALSE}
stainfo <- read.csv("inst/data/stainfo.csv")
glimpse(stainfo)
```

The useable information is containe in the columns `dec_lat_va`, `dec_lon_va`, and `StateCode`. There are services available to lookup timezone given a latitude/longitude pair (e.g. [here](http://www.geonames.org/export/web-services.html), [here](https://developers.google.com/maps/documentation/timezone/intro#Introduction), and [here](https://timezonedb.com/api)), but I'd rather not rely on these. My limited experience has shown them to be slow--only processing one location per request--and anyway I'd prefer to keep this project as self-contained as possible. 

To minimize the number of API calls I'll rely as much as I can on the statecode, and use a simple lookup table, scraped from Wikipedia using the **rvest** package.


This will require a function I've often found handy when splitting strings:

```{r}

# First, a handy function for vectorized splitting of strings
splitPiece <- function (strvec, split, piece, ...) 
{
    spl <- strsplit(strvec, split = split, ...)
    out <- vapply(spl, `[`, character(1), piece)
    out
}

# Now get that table
library(rvest)
library(dplyr)
url <- "https://en.wikipedia.org/wiki/List_of_time_offsets_by_U.S._state"
tzTable <- url %>% 
  read_html(encoding = "UTF-8") %>% 
  html_nodes(xpath = '//*[@id="mw-content-text"]/table[1]') %>% 
  html_table() %>% 
  `[[`(1) %>% 
  setNames(make.names(names(.))) %>% 
  mutate(Time.Offsets = gsub("−", "-", Time.Offsets),
         singleOffset = ifelse(grepl("\n", Time.Offsets), NA, 
                               markstats::splitPiece(Time.Offsets, " ", 1)))
glimpse(tzTable)
```

Now I'll make a table to replace offsets with Olson names (not technically an exact mapping, as Olson names reflect more nuance than offset, but this is a necessary approximation.)

```{r}
unique(tzTable$Time.Offsets)

olsonTbl <- data.frame(offset = sprintf("UTC%03+d:00", c(-10:-5)), 
                       olson = c("Pacific/Honolulu", "America/Anchorage",
                                       "America/Los_Angeles", "America/Denver", 
                                       "America/Chicago", "America/New_York"),
                       stringsAsFactors = FALSE)

kable(olsonTbl)
```

I'll also need a table of FIPS state codes, since that's how state info is given in the siteInfo table. Once again I'll use rvest and Wikipedia:

```{r}
library(xml2)
library(rvest)

scUrl <- "https://en.wikipedia.org/wiki/Federal_Information_Processing_Standard_state_code"

scTable <- scUrl %>% 
  read_html(encoding = "UTF-8") %>% 
  html_nodes(xpath = '//*[@id="mw-content-text"]/table[1]') %>% 
  html_table() %>% 
  `[[`(1) %>% 
  setNames(make.names(names(.))) #%>% 
  # mutate(Time.Offsets = gsub("−", "-", Time.Offsets))

glimpse(scTable)
```


The workflow is then: 

1. Lookup state name using state code
2. Lookup offset using state name
3. Lookup Olson name using offset



```{r}
stateNames <- scTable$Name[match(stainfo$StateCode, scTable$Numeric.code)] %>% 
  gsub("District of Columbia", "Washington, D.C.", .)
stateOffsets <- tzTable$singleOffset[match(stateNames, tzTable$State)]

stateOlson <- olsonTbl$olson[match(stateOffsets, olsonTbl$offset)]

```

Now put this into a function. 


```{r, eval = FALSE}
geonamesTZ <- function(lat, lon, geonamesUser) {
  
  force(geonamesUser)
  
  oneTZ <- function(lat, lon) {
    url <- sprintf("http://api.geonames.org/timezoneJSON?lat=%s&lng=%s&radius=10&username=%s",
                   lat, lon, geonamesUser)
    res <- GET(url)
    stop_for_status(res)
  
    out <- fromJSON(content(res, as = "text", encoding = "UTF-8"))$timezoneID
    if (is.null(out))
      stop("No timezone recognized for location.")
    out
  }
  
  out <- Map(oneTZ, lat = lat, lon = lon)
  out
}


lookupTZ <- function(lat, lon, statecode, geonamesUser = "markwh") {
  data(scTable)
  data(tzTable)
  data(olsonTbl)
  
  state <- scTable$Name[match(statecode, scTable$Numeric.code)]
  state <- gsub("District of Columbia", "Washington, D.C.", state)
  stateOffsets <- tzTable$singleOffset[match(state, tzTable$State)]
  stateOlson <- olsonTbl$olson[match(stateOffsets, olsonTbl$offset)]
  
  if (sum(is.na(stateOlson)) > 0) {
    nas <- is.na(stateOlson)
    stateOlson[nas] <- geonamesTZ(lat = lat, lon = lon, 
                                  geonamesUser = geonamesUser)
  }
  
  stateOlson
}

```

Try it out!

```{r}
lookupTZ(lat = 43, lon = -73, statecode = 1:50)
```
